On the Gravitational stability of gravito-turbulent accretion disks

### 2016-01-25 paper/results2d.tex
finished revising 6.2
version 2 in text, about to redo figs

\section{Two-dimensional disks with beta cooling}\label{result_2d}
We begin in the 2D limit with beta cooling  to make connection 
with previous studies. The disk
material is assumed to be confined to the midplane and $\delta v_z=0$. 
We make the replacement  
$\rho \to \Sigma$, re-interpret $P$ as the vertically-integrated
pressure, and set $\Gamma=1$. % without loss of generality. 
%big gamma only controls vertical structure
%can absorb big gamma into def of cs
The gravitational potential perturbation remains 3D and its midplane
value is given by    
\begin{align}
  \delta \Phi(z=0) = -\frac{2 \pi G}{|k|}\delta\Sigma
\end{align}
\citep{shu70}.

The linearized equations yield an algebraic dispersion relation $s =
s(k)$. We write this  
in terms of the dimensionless growth rate $S = s/\Omega$ and
wavenumber $K=kH = k c_{s0}/\Omega$ as
\begin{align}\label{thindisk}
  f(S,K)\equiv AD - BC = 0,   
\end{align}
where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp},
which do not impose any $\alpha$-$\beta$ relation. We use this generalized
dispersion relation  to investigate GI driven 
by cooling in \S\ref{2d_inviscid}; and GI driven by viscosity 
in \S\ref{2dvisc}.

@@

Eq. \ref{inviscid} is a cubic equation in $S$. The 
Routh-Hurwitz criteria imply that stability is ensured if 
\begin{align}\label{stable_condition}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. 
%gamma/theta condition is related to some kind of thermal instability
%oscilitory convection?? 
%second condition concerning Q is associated with self-g 
Whether or not the system is stable does not 
depend on the cooling time $\beta\in(0,\infty)$, 
but on the irradiation level $\theta$.  
The disk becomes more unstable with decreasing $\theta$, i.e. as the
floor temperature, towards which perturbations are cooled, drops.

If only real growth rates are considered, then violating the second
condition in Eq. \ref{stable_condition} alone is sufficient for instability. 
%from property of cubic crossing the real axis 
In that case the wavenumbers satisfying
\begin{align}
  \theta K^2 - \frac{2}{Q}|K| + 2(2-q) < 0 
\end{align}
are unstable. Thus the range of unstable wavenumbers increases with
decreasing irradiation $\theta$. For $\theta\ll 1$ this range is 
$(2-q)Q\lesssim|K|\lesssim2/\theta Q$. Without irradiation there is no
upper limit to unstable wavenumbers, which could have important
implications for numerical simulations (see \S\ref{prev_works}).

@@

Nevertheless, we can define a characteristic cooling
time $\beta_*$ as that which removes pressure support against
self-gravity over the natural lengthscale in the problem, $H$. We
thus set $|K_*|=1$ and find, for the Keplerian disk,

@@

\subsubsection{$\theta = 1$}\label{theta1}
Beta cooling with $\theta=1$ corresponds to 
`thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case 
$\beta\to 0$ corresponds to an isothermal disk and $\beta
\to \infty$ corresponds to an adiabatic disk. (These two limits are
equivalent for $\gamma=1$.) The most unstable wavenumber and growth
rates vary weakly with $\beta$.

@@

%{\bf maybe move following to discussion}
Numerical simulations of gravito-turbulent disks show there is a
maximum $\alpha$ ($\sim 0.06$) that can be sustained before
fragmentation \citep{rice11}. We can interpret this result in our
linear framework. 
%We can use our case (\ref{gvisc}) to interpret this
Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
by generating a large gravito-turbulent heating rate
($\alpha\gtrsim 1$) through a small $Q$. Fig. \ref{gammie_maxrate_plot} 
shows that such a system would be dynamically unstable with 
growth rate $s = O(\Omega)$. This due to the direct effect of viscous 
stress promoting instability, rather than cooling. Thus, we do not expect rapidly-cooled, 
and hence highly turbulent, self-gravitating disks to persist beyond
dynamical timescales. This is consistent with previous numerical
simulations performed by \cite{lodato05}. 
%Case (\ref{gvisc}) is consistent with 
\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixalpha}\\
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixQ}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.cm
    0.0cm]{figures/result2d_gvisc}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
    1.0cm]{figures/result2d_gvisc_kmax}
  \caption{Growth rates (top) at the optimal wavenumber (bottom) for
    viscous GI as a function of cooling time for the
    case shown in Fig. \ref{gammie_rate_plot}. Solid lines are
    computed from the dispersion relation (Eq. \ref{thindisk}), and the dotted line is an
    estimate from Eq. \ref{gammie_maxrate_simple}. The vertical dashed
    line marks the region with $\alpha > 1$, and the vertical
    dashed-dot line marks the region with growth timescales longer
    than dynamical. 
    \label{gammie_maxrate_plot}}
\end{figure}

>>>>

\section{Two-dimensional disks with beta cooling}\label{result_2d}
We begin in the 2D limit with standard beta cooling  to make connection 
with previous studies. The disk
material is assumed to be confined to the midplane and $\delta v_z=0$. 
We make the replacement  
$\rho \to \Sigma$, re-interpret $P$ as the vertically-integrated
pressure, and set $\Gamma=1$. % without loss of generality. 
%big gamma only controls vertical structure
%can absorb big gamma into def of cs
The gravitational potential perturbation remains 3D and its midplane
value is given by    
\begin{align}
  \delta \Phi(z=0) = -\frac{2 \pi G}{|k|}\delta\Sigma
\end{align}
\citep{shu70}.

The linearized equations yield an algebraic dispersion relation $s =
s(k)$. We write this  
in terms of the dimensionless growth rate $S = s/\Omega$ and
wavenumber $K=kH = k c_{s0}/\Omega$ as
\begin{align}\label{thindisk}
  f(S,K)\equiv AD - BC = 0,   
\end{align}
where the functions $A,B,C,D$ are given in Appendix \ref{2ddisp}. % Note
% that these functions, as given, do not assume any $\alpha$-$\beta$
% relation.
We use this generalized dispersion relation  to investigate GI driven 
by cooling in \S\ref{2d_inviscid}; and GI driven by viscosity 
in \S\ref{2dvisc}.

@@

Eq. \ref{inviscid} is a cubic equation in $S$. The 
Routh-Hurwitz criteria imply that stability is ensured if 
\begin{align}\label{stable_condition}
  \gamma > \theta \quad \text{and} \quad 
  Q > \frac{1}{\sqrt{2\theta(2-q)}} 
\end{align}
are both satisfied. 
%gamma/theta condition is related to some kind of thermal instability
%oscilitory convection?? 
%second condition concerning Q is associated with self-g 
Notice these conditions do not actually
depend on the cooling time. 
The disk becomes more unstable with decreasing $\theta$, 
i.e. as the floor temperature, towards which perturbations are cooled, drops.

If only real growth rates are considered, then violating the second
condition in Eq. \ref{stable_condition} alone is sufficient for instability. 
%from property of cubic crossing the real axis 
In that case the wavenumbers satisfying
\begin{align}
  \theta K^2 - \frac{2}{Q}|K| + 2(2-q) < 0 
\end{align}
are unstable. The range of unstable wavenumbers
increases with decreasing irradiation $\theta$. For $\theta\ll 1$ this range is 
$(2-q)Q\lesssim|K|\lesssim2/\theta Q$. Without irradiation there is no
upper limit to unstable wavenumbers, which could have important
implications for numerical simulations (see \S\ref{prev_works}).

@@

Nevertheless, we can define a characteristic cooling
time $\beta_*$ as that which removes pressure support against
self-gravity over the natural lengthscale in the problem, the
scale-height $H$. We thus set $|K_*|=1$ and find, for the Keplerian
disk,

@@

\subsubsection{$\theta = 1$}\label{theta1}
Standard beta cooling with $\theta=1$ corresponds to 
`thermal-relaxation' in a strictly inviscid disk: the temperature is restored to its initial
value over the cooling time \citep{lin15,mohandas15}. In this case 
$\beta\to 0$ corresponds to an isothermal disk and $\beta
\to \infty$ corresponds to an adiabatic disk. (These two limits are
equivalent for $\gamma=1$.) The most unstable wavenumber and growth
rates vary weakly with $\beta$.

@@

%{\bf maybe move following to discussion}
Numerical simulations of gravito-turbulent disks show there is a
maximum $\alpha$ ($\sim 0.06$) that can be sustained before
fragmentation \citep{rice11}. We can interpret this result in our
linear framework. 
%We can use our case (\ref{gvisc}) to interpret this
Suppose it is possible to balance rapid cooling ($\beta\lesssim1$) 
by generating a large gravito-turbulent heating rate
($\alpha\gtrsim 1$) through a small $Q$. Fig. \ref{gammie_maxrate_plot} 
shows that such a system would be dynamically unstable with 
growth rate $s = O(\Omega)$. This is due to the direct effect of viscous 
stress promoting instability, rather than cooling. Thus, we do not expect rapidly-cooled, 
and hence highly turbulent, self-gravitating disks to persist beyond
dynamical timescales. This is consistent with previous numerical
simulations performed by \cite{lodato05}. 
%Case (\ref{gvisc}) is consistent with 
\begin{figure}
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixalpha}\\
%  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0cm
%    0.0cm]{figures/result2d_fixQ}\\
  \includegraphics[width=\linewidth,clip=true,trim=0cm 1.5cm 0.cm
    0.0cm]{figures/result2d_gvisc}
  \includegraphics[width=\linewidth,clip=true,trim=0cm 0cm 0.cm
    1.0cm]{figures/result2d_gvisc_kmax}
  \caption{Growth rates (top) at the optimal wavenumber (bottom) for
    viscous GI as a function of cooling time for the
    case shown in Fig. \ref{gammie_rate_plot}. Solid lines are
    computed from the dispersion relation (Eq. \ref{thindisk}), and the dotted line is an
    estimate from Eq. \ref{gammie_maxrate_simple}. The vertical dashed
    line marks the region with $\alpha > 1$, and the vertical
    dashed-dot line marks the region with growth timescales longer
    than dynamical. 
    \label{gammie_maxrate_plot}}
\end{figure}

### 2016-01-25 paper/results3d.tex
version 2 in text, about to redo figs

In the following examples we consider a vertically isothermal disk
($\Gamma=1$). Then in the viscous case $\alpha$ is vertically
constant (Eq. \ref{alpha_beta_relation}). We consider even modes about
the midplane, ane apply a numerical disk surface at $z=\zmax$ such
that $\rho(\zmax)=0.05\rho_0$.

@@

We consider an inviscid disk with $\qthree=0.71$ and $\theta=0$. 
Recall the 3D gravity parameter $\qthree$ is defined by 
Eq. \ref{Q3d_def}. This corresponds to a Toomre parameter of
$Q=2Q_\mathrm{crit}$; where $Q_\mathrm{crit}$, defined by
Eq. \ref{qcrit_def}, is the Toomre parameter value such that the 2D
disk would be marginally stable in the absence of cooling.

>>>>

In the following examples we consider a vertically isothermal disk
($\Gamma=1$). Then in the viscous case $\alpha$ is vertically
constant (Eq. \ref{alpha_beta_relation}). We consider even modes about
the midplane, and apply a numerical disk surface at $z=\zmax$ such
that $\rho(\zmax)=0.05\rho_0$.

@@

We consider a 3D inviscid disk with $\qthree=0.71$ and $\theta=0$. 
The 3D gravity parameter $\qthree$ is defined by 
Eq. \ref{Q3d_def}. For this case the corresponding Toomre parameter 
$Q=2Q_\mathrm{crit}$. Recall $Q_\mathrm{crit}$, defined by
Eq. \ref{qcrit_def}, is the Toomre parameter value such that the 2D
disk would be marginally stable in the absence of cooling.

### 2016-01-25 paper/summary.tex
version 2 in text, about to redo figs

\section{Summary and discussion}\label{summary}
In this paper, we develop the linear theory of cooling, irradiated,
and viscous accretion disks in order to understand gravitational
instability (GI) in realistic protoplanetary disks (PPDs). 
We use a Navier-Stokes viscosity to mimic the effects of turbulence. 
This viscosity provides a background heating to balance the imposed cooling,  
and may also act on linear perturbations. %thermo and dynamic 
We suggest that disk fragmentation, observed in numerical simulations, 
can be understood as the eventual outcome of secondary
instabilities of the gravito-turbulent base state, driven by cooling
and/or viscosity.     
%This %allows us to assess realistic disk fragmentation without using
%uncertain empirical conditions based on numerical simulations.   
%{\bf cooling removes pressure support, viscosity removes rotational support}
%{\bf our conclusions do require input of emperical resuls from numerical
%sims}
%going beyond toomre 
%move to subsection?
While viscosity and cooling can be related through thermal
balance, they independently enhance growth rates: cooling reduces
thermal stabilization; and viscous forces compromise rotational
stabilization. This provides a physical explanation as to why
rapidly-cooled, gravito-turbulent disks cannot exist.

@@

The effect of cooling and irradiation on GI is quantified by the 
dispersion relation 
Eq. \ref{inviscid}. The sufficient condition for stability 
depends on the irradiation level (Eq. \ref{stable_condition}), and is 
\emph{independent} of the cooling time, provided it is non-zero and
finite. This means that long cooling times can still  
formally lead to instability. However, growth timescales may be 
uninterestingly long for cooling times $\tcool\Omega\gtrsim 
O(10)$. Because cooling affects pressure support, GI driven by cooling
occur on small scales, $kH\gtrsim O(1)$.

@@

We generalize the `viscous gravitational instability', previously 
studied in isothermal disks
\citep{lynden-bell74,willerding92,gammie96}, to include 
cooling, viscous heating and irradiation. In our fiducial disk,  
where viscosity and self-gravity are inversely
related ($\alpha\propto Q^{-2}$) to model a
gravito-turbulent background, this instability occurs on orbital  
timescales for $\alpha\gtrsim 0.1$. This is consistent with the notion
of a maximum stress sustainable by gravito-turbulence established
through numerical simulations \citep{rice05}. Because viscosity affects 
rotational support, viscous GI occurs on large scales, $kH\lesssim
O(1)$.  Furthermore, irradiation preferentially
stabilizes small-scale perturbations. %also true for inviscid case
%maybe that's why irradiated disks produce smaller alpha

@@

We apply our linear framework to protoplanetary disks with 
realistic models for cooling and gravito-turbulence. 
%\citep{rafikov15}. 
%We show that, for linear dynamics, realistic 
%cooling in a physically non-irradiated disk ($T_\mathrm{irr}=0$) is
%actually modeled by `beta' cooling with a  
%non-zero irradiation parameter ($\theta\neq0$), in contrast to 
%numerical simulations which typically employ beta cooling with  
%$\theta=0$ \citep[e.g.][]{gammie01}. How relevant 
%are such simulations to realistic PPDs must then be further 
%investigated. 
%The important implication of having 
%$\theta\sim 1$ in a PPD is that cooling alone cannot 
%destabilize the disk at large radii. 
We show that with a realistic cooling model for PPDs,
cooling alone cannot lead to gravitational instabilities. This is due
to stabilization by an effective irradiation associated with the PPD
cooling function, which is present even if 
there is no physical irradiation. %due to matching functional forms of
                                %cooling 
Instead, viscous GI occurs on  
dynamical timescales for $R\gtrsim 60$AU because 
$\alpha\gtrsim0.1$ there. This corresponds to a Toomre $Q\simeq 1.5$ and 
a cooling time $\tcool\lesssim 3\Omega^{-1}$. These are coincident with 
\emph{empirical} conditions cited in the literature to determine disk
fragmentation 
\citep[e.g.][]{rafikov15}. Here, we attribute a \emph{physical}
cause for the fragmentation of realistic PPDs: gravito-turbulent PPDs
fragment when turbulent stresses are large enough to further
destabilize the disk against self-gravity.    
%collary: details of cooling unimportant as long as the associated
%stress is the same

@@

%we do not look at fragment survival
%fragment due to cooling or viscosity?
%they should produce diff size frags
\subsection{Relation to numerical simulations}\label{prev_works}
%frag in sims because of : cooling, viscosity, or high den
%flucuations? don't know 
Our results may shed light on some numerical simulations 
of disk fragmentation. Table 
\ref{bstar_compare} shows a close match between the characteristic
cooling time for cooling-driven GI (Eq. \ref{betastar}) 
and that for disk fragmentation observed in simulations 
\citep{gammie01,rice05,rice11}. This suggests that, at least for those
simulations, fragmentation is physically due to the removal of
pressure stabilization by cooling on a radial scale of the disk scale
height. 
%i.e. cool-driven g 
{\bf don't know if we want to dwell on this table}

More recent simulations have raised the issue of numerical
convergence. \cite{paardekooper12} found that fragmentation can occur
for   
slowly-cooled disks, $\tcool\Omega\gg O(1)$, but that this requires   
simulations to run for significantly longer than dynamical 
timescales. This is consistent with our finding
that for either cooling-driven or viscous GI, there is no critical 
cooling rate/viscosity below  which the disk is  
formally stable. Instead, growth rates smoothly decrease with
increasing $\tcool$ (decreasing $\alpha$), implying that
instabilities, and hence fragmentation, simply take longer to develop
for slowly-cooled disks.         
%hopkins paper

@@

We thus suggest that having a floor temperature in the standard beta
cooling prescription is a necessary condition for  
convergence with respect to numerical resolution. This limits the
relevant scales to a finite range.  %for cool gi 
%physically irradiation 
Furthermore, without a floor temperature, standard beta cooling only
depends on the pressure/energy density field. It is not clear how
results obtained from such models can be applied to PPDs, where
cooling depends on two thermodynamic states (e.g. $P$ and
$T$). A floor temperature permits a mapping between beta
cooling and realistic cooling (\S\ref{ppd_cooling}).

@@

On the other hand, \cite{baehr15} included a floor temperature (or irradiation)
in their simulations but still find that at fixed cooling rates, disks
eventually fragment with sufficient spatial resolution.
In light of our results on viscous GI, 
we here highlight a possible contribution: %in addition to num visc (meru lodato)
high resolutions enable small-scale turbulent angular momentum transport 
% physical, cf numerical  
to aid clump formation (see also \S\ref{MHD} below). 
This mechanism is unavailable at low resolutions, for which one must
rely on rapid cooling.

%This, in fact, has been suggested for

We comment that although modern simulations resolve the dominant
scale associated with gravito-turbulence, $kH\sim 1$
\citep{cossins09}, this does not
necessarily imply small-scale dynamics/thermodynamics are unimportant 
\citep[especially for non-linear evolution, see][]{young15},  
as is evident from the non-converging simulations described above.

@@

\subsection{MHD turbulence}\label{MHD}%turbulence and viscosity 
We emphasize the linear theory we have developed does not assume  
a particular origin for the turbulence represented by the imposed
viscosity. For example, the 2D dispersion relation, Eq. \ref{thindisk}
and Appendix \ref{2ddisp}, treats $\alpha$ as an independent
input parameter.

@@

%effect of MHD turbulence on GI 
Our models may thus apply to self-gravitating disks dominated by MHD
turbulence.     
%In fact, the viscous approach may be more
%applicable here than gravito-turbulence, because the characteristic
%wavelength of the magneto-rotational instability (which leads to MHD
%turbulence) is $\ll H$, at least in the ideal MHD limit. Then there is
%a scale separation between the underlying, small-scale MHD 
%turbulence and the large-scale effects of self-gravity. 
%fromang is isothermal sims - no cooling effect 
Explicit numerical simulations of magnetized, massive disks have been performed 
by \cite{fromang05}. He finds disk fragmentation with increasing numerical
resolution, attributing this to resolving the most unstable MRI
wavelength, which enables small-scale angular momentum removal by MHD
turbulence to aid fragmentation. %this supports using viscosity to
                                %mimic turbulence
This would be a physical mechnism represented by the viscous GI
discussed in this paper; which lends some support for  
the use of a viscosity to represent turbulence.

@@

\subsection{Outstanding issues}
%the link between the linear GI's discussed in this paper 
%and frag 
True disk fragmentation is a non-linear process characterized by
clumps reaching densities orders of magnitude above the ambient value, and
which survives tidal shear and shocks \citep{paardekooper12,young16}.  
%cite the young papers 
Clearly, our linear models cannot address fragmentation
directly. Technically, we have only demonstrated that a 
gravito-turbulent state becomes dynamically unstable, and thus should
not persist, when cooling is too rapid or when the associated viscous
stresses are too large.  However, numerical simulations show that
there are only two possible outcomes of cooling, self-gravitating
disks: steady gravito-turbulence or fragmentation. It seems then
reasonable to  speculate that the non-existence of a stable
gravito-turbulent state, here due to dynamical instability, implies
disk fragmentation.

@@

Our deterministic approach cannot model `stochastic fragmentation'
\citep{paardekooper12,hopkins13}. In this interpretation,  
fragmentation is attributed to the occurence (and survival) 
of non-linear density perturbations, which arise from the
gravito-turbulent state simply by chance. 
%Whether these high
%density perturbations actually lead to a clump formation, 
This might be considered as a secondary instability by lowering the
Toomre $Q$ through a (random) increase in density
(Eq. \ref{toomreQ_criterion}). By contrast, the viscosity/cooling
mechanisms discussed in this paper lowers $Q$ by reducing pressure and
Coriolis forces. %we have a physical way of lowering pressure/rotation

@@

%modeling turb as visc %explore other visc laws 
The most important assumption in this work is modeling turbulence as a  
Navier-Stokes viscosity. Furthermore, we have chosen a particular  
viscosity law (see \S\ref{visc_model}) to mimic the effect of 
turbulence in reducing rotational support (in the sense that it
provides small-scale angular momentum transport). 
%to enhance clumping 
%this visc law is used in prev studies (by much smarter people)
How to quantitatively model turbulence (be it due to self-gravity or
MHD) as a viscosity, especially on dynamical timescales, should be
clarified with direct numerical simulations. The present linear models
should then be modified accordingly. 
%could consider other visc parameters mu and lambda (but this
%introduces visc overstability, or new functional forms of the
%viscosity law altogether

>>>>

\section{Summary and discussion}\label{summary}
In this paper, we develop the linear theory of cooling, irradiated,
and viscous accretion disks in order to understand gravitational
instability (GI) in realistic protoplanetary disks (PPDs). 
We use a Navier-Stokes viscosity to mimic the effects of
turbulence. %which may be of any origin, though we specialize to
            %gravito-turb 
This viscosity provides a background heating to balance the imposed cooling,  
and may also act on linear perturbations. %thermo and dynamic 
We suggest that disk fragmentation, observed in numerical simulations, 
can be understood as the eventual outcome of secondary
instabilities of a gravito-turbulent base state, driven by cooling
and/or viscosity.     
While viscosity and cooling can be related through thermal
balance, they independently enhance growth rates: cooling reduces
thermal stabilization; and viscous forces compromise rotational
stabilization. This provides a physical explanation as to why
rapidly-cooled, gravito-turbulent disks cannot exist.

@@

The effect of cooling and irradiation on GI is quantified by the 
dispersion relation 
Eq. \ref{inviscid}. We find a sufficient condition for stability which  
depends on the irradiation level (Eq. \ref{stable_condition}) but 
\emph{independent} of the cooling time. 
This means that long cooling times can still  
formally lead to instability. However, growth timescales may be 
uninterestingly long for cooling times $\tcool\Omega\gtrsim 
O(10)$. Because cooling affects pressure support, GI driven by cooling
occur on small scales, $kH\gtrsim O(1)$.

@@

We generalize the `viscous gravitational instability', previously 
studied in isothermal disks
\citep{lynden-bell74,willerding92,gammie96}, to include 
cooling, viscous heating and irradiation. %and 3D 
We consider a disk with viscosity and self-gravity inversely
related ($\alpha\propto Q^{-2}$) to model a 
gravito-turbulent background, and find viscous GI occurs on orbital
timescales for $\alpha\gtrsim 0.1$. This is consistent with the notion 
of a maximum stress sustainable by gravito-turbulence established 
by numerical simulations \citep{rice05}. Because viscosity affects 
rotational support, viscous GI occurs on large scales, $kH\lesssim
O(1)$.  Furthermore, irradiation preferentially
stabilizes small-scale perturbations. %also true for inviscid case
%maybe that's why irradiated disks produce smaller alpha

@@

We apply our linear framework to protoplanetary disks with 
realistic models for cooling and gravito-turbulence. 
We show that with a realistic cooling model for PPDs, 
cooling alone cannot lead to gravitational instabilities. This is due
to stabilization by an effective irradiation associated with the PPD
cooling function, which is present even if 
there is no physical irradiation. %due to matching functional forms of
                                %cooling 
Instead, viscous GI occurs on  
dynamical timescales for $R\gtrsim 60$AU because 
$\alpha\gtrsim0.1$ there. This corresponds to a Toomre $Q\simeq 1.5$ and 
a cooling time $\tcool\lesssim 3\Omega^{-1}$. These are coincident with 
\emph{empirical} conditions cited in the literature to determine disk
fragmentation 
\citep[e.g.][]{rafikov15}. Here, we attribute a \emph{physical}
cause for the fragmentation of realistic PPDs: gravito-turbulent PPDs
fragment when turbulent stresses are large enough to further
destabilize the disk against self-gravity.    
%collary: details of cooling unimportant as long as the associated
%stress is the same

@@

\subsection{Relation to numerical simulations}\label{prev_works}
%frag in sims because of : cooling, viscosity, or high den
%flucuations? don't know 
Our results may shed light on some numerical simulations 
concerning disk fragmentation. Table 
\ref{bstar_compare} shows a close match between the characteristic
cooling time for cooling-driven GI (Eq. \ref{betastar}) 
and that for disk fragmentation observed in simulations 
\citep{gammie01,rice05,rice11}. This suggests that, at least for those
simulations, fragmentation is physically due to the removal of
pressure stabilization by cooling on a radial scale of the disk scale
height. %i.e. cool-driven g 
{\bf don't know if we want to dwell on this table}
%negligible effect from viscous heating
%makes sense? sims show clumps are not turbulent internally
%then visc plays no role in clumping -> only a cooling effect

More recent simulations have raised the issue of numerical
convergence. \cite{paardekooper12} found that fragmentation can occur 
for  slowly-cooled disks with $\tcool\Omega\gg O(1)$, but that this requires   
simulations to run for significantly longer than dynamical 
timescales. This is consistent with our finding
that for either cooling-driven or viscous GI, there is no critical 
cooling rate/viscosity below  which the disk is  
formally stable. Instead, growth rates smoothly decrease with
increasing $\tcool$ (decreasing $\alpha$), implying that
instabilities, and hence fragmentation, simply take longer to develop
for slowly-cooled disks.         
%hopkins paper

@@

We suggest having a floor temperature in the standard beta
cooling prescription is a necessary condition for  
numerical convergence. This limits the
relevant scales to a finite range.  %for cool gi 
%physically irradiation 
Furthermore, without a floor temperature, standard beta cooling is
a function of the pressure/energy density field only. It is not clear
how relevant are results obtained from this is to actual PPDs,
where cooling depends on two thermodynamic states (e.g. pressure and
temperature). A floor temperature permits a mapping between standard beta
cooling and PPD cooling (\S\ref{ppd_cooling}).

@@

On the other hand, \cite{baehr15} included a floor temperature 
in their simulations but still find that at fixed cooling rates, disks
eventually fragment with sufficient spatial resolution.
In light of our results on viscous GI, 
we here highlight a possible contribution: %in addition to num visc (meru lodato)
high resolutions enable small-scale turbulent angular momentum transport 
% physical, cf numerical  
to aid clump formation (see also \S\ref{MHD} below). 
This mechanism is unavailable at low resolutions, for which one must
rely on rapid cooling.  
%maybe there is an effective floor temp which depends on res

We comment that although modern simulations resolve the dominant
scale associated with gravito-turbulence, $kH\sim 1$
\citep{cossins09}, this does not
necessarily imply small-scale dynamics/thermodynamics are unimportant 
\citep[especially for non-linear evolution, ][]{young15},  
as is evident from the non-converging simulations described above.

@@

\subsection{MHD turbulence}\label{MHD}%turbulence and viscosity 
We emphasize the linear framework we have developed does not assume  
a particular origin for the turbulence that is represented by the imposed
viscosity. For example, the 2D dispersion relation given by Eq. \ref{thindisk}
with Eq. \ref{bigA}---\ref{bigF}, treats $\alpha$ as an independent
input parameter.

@@

%effect of MHD turbulence on GI 
Our models may thus apply to self-gravitating disks dominated by MHD
turbulence.     
%In fact, the viscous approach may be more
%applicable here than gravito-turbulence, because the characteristic
%wavelength of the magneto-rotational instability (which leads to MHD
%turbulence) is $\ll H$, at least in the ideal MHD limit. Then there is
%a scale separation between the underlying, small-scale MHD 
%turbulence and the large-scale effects of self-gravity. 
%fromang is isothermal sims - no cooling effect 
Explicit numerical simulations of magnetized, massive disks have been performed 
by \cite{fromang05}. He finds disk fragmentation with increasing numerical
resolution, and attributes this to resolving the most unstable MRI
wavelength, which enables small-scale angular momentum removal by MHD
turbulence to aid fragmentation. %clumping/mass accretion onto clump
This would be a physical mechnism represented by the viscous GI 
discussed in this paper; which lends some support for  
the use of a viscosity to represent turbulence.

@@

\subsection{Outstanding issues}
%the link between the linear GI's discussed in this paper 
%and frag 
True disk fragmentation is a non-linear process characterized by
clumps reaching densities orders of magnitude above the ambient value, and
which survives tidal shear and shocks \citep{paardekooper12,young16}.  
Clearly, our linear models cannot address fragmentation
directly. Technically, we have only demonstrated that a 
gravito-turbulent state becomes dynamically unstable, and thus should
not persist, when cooling is too rapid or when the associated viscous
stresses are too large.  However, steady gravito-turbulence and
fragmentation are the two \emph{only} possible outcomes of cooling, 
self-gravitating disks in the local limit \citep[][]{gammie01}.    
Then it seems reasonable to speculate that the non-existence of a stable
gravito-turbulent state, here due to dynamical instability, implies
disk fragmentation.

@@

Our deterministic approach cannot model `stochastic fragmentation'
\citep{paardekooper12,hopkins13}. In this interpretation,  
fragmentation is attributed to the occurence and survival 
of large, non-linear density enhancements, which arise from the
gravito-turbulent flucuations simply by chance. 
%cooling affects survival rate 
This might be considered as a secondary instability by lowering the
Toomre parameter through a (random) increase in density 
(Eq. \ref{toomreQ_criterion}). By contrast, the viscosity/cooling
destabilizing mechanisms discussed in this paper lowers $Q$ by
reducing pressure and Coriolis forces. %we have a physical way of lowering pressure/rotation

@@

%modeling turb as visc %explore other visc laws 
The most important assumption in this work is modeling turbulence as a  
Navier-Stokes viscosity. Furthermore, we have chosen a particular  
viscosity law (see \S\ref{visc_model}) to mimic the effect of 
turbulence in reducing rotational support (in the sense that it
provides small-scale angular momentum transport). 
%to enhance clumping 
%this visc law is used in prev studies (by much smarter people)
How to quantitatively model gravito- or MHD turbulence 
as a viscosity, especially on dynamical timescales, should be 
clarified with direct numerical simulations. The present viscosity
models should then be modified accordingly. 
%could consider other visc parameters mu and lambda (but this
%introduces visc overstability, or new functional forms of the
%viscosity law altogether
